National Pipeline System as a Solution to Future Water Shortage in the United States

### 2013-02-04 paper.tex
Title added
lol

\title{title}

@@

This paper attempts to address key aspects of a feasible water management
strategy, based on the available data. With our projection of expected
water needs in mind, we aim to meet demands by investing in the following
infrastructure: 
\begin{itemize}
\item De-salinization plants 
\item Water purification facilities 
\item Reservoirs and dams
\end{itemize}
However, it is not sufficient to simply have more water available
in the country's supplies as a whole. It's also critical that each
individual location be able to access that water. For this purpose
we aim to find the optimal locations to place pipelines.

This paper is  concerned with the construction of new water
supply infrastructure. Thus, there are some aspects of the water management
process that will not be further considered: 
\begin{itemize}
\item Public education: As with all issues of resource conservation, the
behaviour of the general public has an effect on water supplies. This
will not be considered here, for several reasons. The key issue is
that there is a lack of empirical evidence suggesting that currently
known public education strategies will work to minimise water usage
in the United States (developing such strategies further is beyond
the scope of this paper). The United States in particular has been
particularly resistant to attempts to reduce per capita water consumption\cite{key-7}.
More generally, even water education programmes considered particularly
successful, such as that of Singapore, have not managed to make a
huge impact on consumption, with an approximate 5 per cent reduction
in household water use over 6 years\cite{key-8}. Finally, the effect
of water education programmes is likely to have a similar effect in
each location. This means that areas that would be most in need of
additional infrastructure taking into account such programmes would
be essentially the same as those areas most in need while not taking
into account such programmes. 
\item Maintenance costs of current infrastructure: minimizing maintenance
costs depends on technological advancement and management, but is
essentially a fixed cost irrespective of the new infrastructure we
decide to implement.
\item Cost of obtaining data: Much of our analysis depends on USGS data.
We note that the cost of obtaining this survey data is negligible
(the WaterSMART initiative, which collects much of the data, cost
\$21 million this year\cite{key-9}. In comparison, de-salinization
facilities can easily cost billions to construct\cite{key-10}) compared
to the utility of superior decision making in constructing water-related
infrastructure (among various other things the data is useful for).
\end{itemize}

@@

Due to the scale and complexity of the task of water management in
the United States, we make several simplifying assumptions in order
to obtain a workable model.
\begin{itemize}
\item Counties are thought of as discrete entities, occupying points in
space located at their centroid. This assumption enables us to think
of counties as vertices (as in a graph). This assumption can be made
without losing too much accuracy for several reasons: The United States
is split into more than 3,000 counties, with an average area of less
than 3,300 square kilometers, and average length across of less than
60 kilometers. The level of inaccuracy is sufficiently low that it
is workable (in contrast, states for instance are far too large to
be accurately modeled as points).
\item When considering the costs of pipelines, only the distance of pipeline
is considered. The landscape of a region may in practice affect the
length of pipeline required, or may result in additional cost per
length of pipeline, but we shall not model this, as we could not find
data on the difficulty of pipeline construction across different types
of terrain.
\item Pipelines are thought of as bi-directional. This assumption essentially
reflects the reality of pipelines. There are a few exceptional cases
where it is significantly more difficult to send water in one direction,
but these are rare enough that they can be disregarded.
\item Pipelines have variable rates of water transmission, depending on
factors such as size. In theory it would be possible to save money on pipelines by constructing smaller pipelines in areas where water rarely needs to be transported at high rates. 
However, the amount of water that needs to be transported through any given pipe is highly variable (ie. after a sudden intense rainfall). Thus, pipelines for long-distance water transmissions are relatively stable in cost. Our estimate below for pipeline cost is a typical figure.
\item Some pipelines exist already, but we do not consider them in our model. This is partially because they are likely to have minimal impact because there are not many of them, and partially because already-existing pipelines are poorly documented, so there is no way to efficiently collect and parse such data
\end{itemize}

@@

We assume a constant cost per length of pipeline. Estimates for the
cost of pipeline range from approximately \$1.5 million per mile to
\$5 million per mile\cite{key-11}, with the higher figures generally
corresponding to older pipelines constructed less efficiently than
is possible today. We assume a figure of \$2 million per mile for
the purporses of our pricing estimates.

@@

\section{Optimising Placement of Pipelines}

@@

To optimise the placement of pipelines, we parse geographical data
for each of the counties, and produce a graph in which counties are
represented as vertices, and undirected edges are placed between adjacent
counties, with weights corresponding to the distance between the counties
being connected.

@@

To do this, we implement Kruskal's minimum spannning tree algorithm
on the complete graph of counties with edges weighted by distance.

@@

To an extent

The cost of the proposed infrastructure developments are significant,
but must be placed in context. The cost of the 2012-2013 drought season
on the United States economy is estimated at between \$75 billion
and \$150 billion, and the potential total cost to the economy of
drought-related occurrences between 2013 and 2025 is much greater
even still.

>>>>

\title{National Pipeline System as a Solution to Future Water Shortage in the United States}

@@

This paper attempts to address key aspects of a feasible water management
strategy, based on the available data. With our projection of expected
water needs in mind, we aim to meet demands by investing in the following
infrastructure: 
\begin{itemize}
\item De-salinization plants 
\item Water purification facilities 
\item Reservoirs and dams
\end{itemize}
However, it is not sufficient to simply have more water available
in the country's supplies as a whole. It's also critical that each
individual location be able to access that water.

 For this purpose
we aim to find the optimal locations to place pipelines.

This paper is  concerned with the construction of new water
supply infrastructure. Thus, there are some aspects of the water management
process that will not be further considered: 
\begin{itemize}
\item Public education: As with all issues of resource conservation, the
behaviour of the general public has an effect on water supplies. This
will not be considered here, for several reasons. The key issue is
that there is a lack of empirical evidence suggesting that currently
known public education strategies will work to minimize water usage
in the United States (developing such strategies further is beyond
the scope of this paper). The United States in particular has been
particularly resistant to attempts to reduce per capita water consumption\cite{key-7}.
More generally, even water education programmes considered particularly
successful, such as that of Singapore, have not managed to make a
huge impact on consumption, with an approximate 5 per cent reduction
in household water use over 6 years\cite{key-8}. Finally, 
water education programmes are likely to have a similar effects in
each location — there is no particular reason to believe they would be much more successful in any one location. This means that areas that would be most in need of
additional infrastructure taking into account such programmes would
be essentially the same as those areas most in need while not taking
into account such programmes. 
\item Maintenance costs of current infrastructure: minimizing maintenance
costs depends on technological advancement and management, but is
essentially a fixed cost irrespective of the new infrastructure we
decide to implement.
\item Cost of obtaining data: Much of our analysis depends on USGS data.
We note that the cost of obtaining this survey data is negligible
(the WaterSMART initiative, which collects much of the data, cost
\$21 million this year\cite{key-9}. In comparison, de-salinization
facilities can easily cost billions to construct\cite{key-10}) compared
to the utility of superior decision making in constructing water-related
infrastructure (among various other things the data is useful for).
\end{itemize}

@@

Due to the scale and complexity of the task of water management in
the United States, we make several simplifying assumptions in order
to obtain a workable model.
\begin{itemize}
\item Counties are thought of as discrete entities, occupying points in
space located at their centroid. This assumption enables us to think
of counties as vertices (as in a graph). This assumption can be made
without losing too much accuracy for several reasons: The United States
is split into more than 3,000 counties, with an average area of less
than 3,300 square kilometers, and average length across of less than
60 kilometers. The level of inaccuracy is sufficiently low that it
is workable (in contrast, states for instance are far too large to
be accurately modeled as points).
\item De-salinization plants, reservoirs, and purification plants are all facilities that are built at a single location, and have the intended effect of providing more water to that location. For the purposes of our modelling, we think of them as being essentially the same kind of thing.
\item When considering the costs of pipelines, only the distance of pipeline
is considered. The landscape of a region may in practice affect the
length of pipeline required, or may result in additional cost per
length of pipeline, but we shall not model this, as we could not find
data on the difficulty of pipeline construction across different types
of terrain.
\item Pipelines are thought of as bi-directional. This assumption essentially
reflects the reality of pipelines. There are a few exceptional cases
where it is significantly more difficult to send water in one direction,
but these are rare enough that they can be disregarded.
\item Pipelines have variable rates of water transmission, depending on
factors such as size. In theory it would be possible to save money on pipelines by constructing smaller pipelines in areas where water rarely needs to be transported at high rates. 
However, the amount of water that needs to be transported through any given pipe is highly variable (ie. after a sudden intense rainfall). Thus, pipelines for long-distance water transmissions are relatively stable in cost. Our estimate below for pipeline cost is a typical figure.
\item Some pipelines exist already, but we do not consider them in our model. This is partially because they are likely to have minimal impact because there are not many of them, and partially because already-existing pipelines are poorly documented, so there is no way to efficiently collect and parse such data
\end{itemize}

@@

We assume a constant cost per length of pipeline. Estimates for the
cost of pipeline range from approximately \$1.5 million per mile to
\$5 million per mile\cite{key-11}, with the higher figures generally
corresponding to older pipelines constructed less efficiently than
is possible today. We assume a figure of \$2 million per mile for
the purposes of our pricing estimates.

@@

\section{Optimizing Placement of Pipelines}

@@

To optimize the placement of pipelines, we parse geographical data
for each of the counties, and produce a graph in which counties are
represented as vertices, and undirected edges are placed between adjacent
counties, with weights corresponding to the distance between the counties
being connected.

@@

To do this, we implement Kruskal's minimum spanning tree algorithm
on the complete graph of counties with edges weighted by distance.

@@

The cost of the proposed infrastructure developments are significant, at a total of $171 billion,
but must be placed in context. The cost of the 2012-2013 drought season
on the United States economy is estimated at between \$75 billion
and \$150 billion, and the potential total cost to the economy of
drought-related occurrences between 2013 and 2025 is much greater
even still.
